Colonies - Compute Continuums across Platforms

### 2023-05-03 paper_2023/paper.tex
wip

\begin{abstract}
Artificial intelligence and machine learning has gained significant traction in recent years. At the same time, development and operation of AI workloads has become increasingly challenging. One difficulty is the lack of portability, making it cumbersome to move from one platform to another. Creating and operating fully automated end-to-end workflows across devices, edge, and cloud platforms is even more challenging.

@@

Based on a stateless microservice architecture, complex workflows can be broken down into composable functions. These composable functions can be implemented in any programming language and be executed by distributed executors deployed across platforms, anywhere on the Internet such as IoT, cloud, edge, or a supercomputer. A zero-trust security protocol enables a collection of distributed executors to operate as a unified entity, establishing seamless compute continuums across platforms.

@@

Taking an AI model into production requires substantial software engineering expertise and collaboration across teams. In contrast to traditional IT workloads, both the data and the model must be managed in addition to the software itself. As most models require regular retraining or re-calibration, it must be possible to update deployed models and software seamlessly without losing information or breaking the system. In many cases, there is a constant flow of data ingested into the system which must be managed even in case of failures. This becomes even more challenging when nodes or parts of the underlying infrastructure become unavailable due to maintenance such as software updates, hardware replacements and sometimes misconfiguration problems.

In some cases, it may be necessary to scale the system to increase or reduce the capacity. This is especially critical when using expensive cloud resources. Scaling the system means that the underlying infrastructure may change at any time, causing instability issues for running services or workflows. Therefore, it must be possible to detect failed computations and reprocess failed tasks part of a larger workflow. Workflows must hence be designed to handle an ever-changing infrastructure, and if a failed computation cannot be restored gracefully, engineers must be able to quickly perform root cause analysis to manually recover the system.

@@

Additionally, many compute clusters operate on-premises installations. Sometimes it is necessary to temporarily increase the capacity of on-prem clusters by combining resources from multiple providers, for example, adding cloud compute resources to handle peak loads or utilize HPC resources to quickly reprocess historical data. Developing hybrid workflows where some jobs run in the cloud and others run on HPC systems requires even more software development efforts \cite{wf_challenges} and is beyond the scope of many users, preventing them from utilizing powerful hardware. Clearly, there is a need for a framework that can consolidate various workflow management platforms to simplify development and enable seamless execution across platforms.

The long-term vision of the Colonies framework is to enable uninterrupted access to computing resources, allowing data to flow from one service to another establishing a seamlessly integrated computing ecosystem. To realize this vision, the paper presents \emph{Colonies}, a \emph{distributed runtime environment}. The remainder of the paper describes the Colonies framework and how it can be used to create robust and scalable compute continuums.

@@

Today, utilization of serverless computing is experiencing a significant growth \cite{cognit}, primarily due to its potential to liberate developers from the burden of managing underlying cloud infrastructures. Attempts are currently being made to implement serverless workflow management systems. For example, OpenWolf \cite{openwolf} is a serverless workflow engine designed to utilize the Function-as-a-Service (FaaS) paradigm for composing complex scientific workflows. It is based on OpenFaaS \cite{openfaas}, which allows functions to run on Kubernetes clusters.

Colonies can also be used to implement serverless workflow management systems. This will be further discussed in Section \ref{faas}. However, in contrast to previous work, Colonies is designed to be platform independent and does therefore not require Kubernetes. By using a zero-trust security protocol, functions can be securely executed by distributed executors deployed anywhere on the Internet. It is important to point out that Colonies does not provide an infrastructure for function execution. Instead, Colonies primary role is to serve as a platform for coordinating function executions which are carried out by distributed executors.

Currently, microservices are primarily used to implement large-scale web applications or Internet applications requiring high-availability. It has not yet become a prevalent design principle for workload management or implementation of HPC applications. Instead, simple job scripts are commonly used. J. Represa et al. \cite{GarciaRepresa1740746, GarciaRepresa1640771} explore various challenges associated with developing microservice-based workflow management for industrial automation within the context of the Arrowhead project \cite{delsing2017iot}. The authors conclude that microservice-based workflow technologies are viable for industrial applications, particularly due to their inherent flexibility. The primary contribution of this paper is a comprehensive technical description of how to implement a distributed workflow engine based on microservice principles, extending beyond orchestrating microservices to execute automation tasks within one platform. The vision is to create a distributed architecture where microservices can reside anywhere on the Internet and still function as a cohesive unit to execute cross-platform workloads.  
o

@@

\section{The Colonies framework}
\label{sec:headings}
A core concept of the Colonies framework is the notion of \emph{processes}. A process contains meta-information about computations executed by remote computer programs, referred to as \emph{distributed executors}. Specifically, a process contains a definition of a function to be executed as well as contextual information such as execution status, including the result of the computation. One can think of a process as a digital twin of a real computing process. It is essential to note that a process does not necessarily have to be an operating system process, but can represent any type of computation, e.g. a remote procedure call executed by any kind of software.

@@

\emph{Distributed executors} are responsible for executing processes assigned by the Colonies Servers. Several distributed executors form a \emph{Colony}, which is a collection of executors operating as a cohesive unit where each executor is responsible for executing specific types of processes. One could view a Colony as an organization or a society of distributed computer softwares acting as a single virtual computer system. To interact with other executors, executors must prove their Colony membership using a cryptographic protocol that follows the zero-trust security principle: \emph{never trust, always verify}. This security protocol ensures that users can keep control even when executors are distributed across platforms. Zero-trust security is fundamental to the Colonies framework and will be further discussed in Section \ref{zerotrustsecurity}. The following section outlines the underlying design principles and describe the Colonies framework in more detail.

@@

\begin{itemize}
\item \textbf{Single responsibility:} Each executor is only accountable for executing specific functions. This makes the system easier to understand, develop, test and maintain.  
\item \textbf{Loosely coupled:} Executors are designed with minimal dependencies on other executors, enabling various software development teams to work independently. For instance, a data engineering team may handle the implementation of pre-processing functions, while a data science team oversees machine learning functions, and another team manages visualization or customer integration etc.

\item \textbf{Scalability:} Executors can be deployed independently, which enables horizontal scaling. This allows for better resource utilization, parallelism, and improved performance.

\item \textbf{Resilience:} The failure of a single executor does not compromise the entire application or workflow. Executors' isolation from one another contributes to a more resilient and fault-tolerant system. If an executor crashes during execution, the process is automatically reassigned to another executor.

@@

\item \textbf{Technology agnostic:} Executors can be implemented in any programming language, facilitating seamless integration with other applications and systems.

\item \textbf{Decentralized governance:} As the Colonies framework is technology-agnostic, different software development teams can make independent technology and design decisions when developing executors, promoting greater flexibility and adaptability. For example, some executors may be implemented in Rust, while others may use Python to leverage state-of-the-art machine learning frameworks.
\end{itemize}

Although microservices have attractive properties and can simplify both development and deployment of executors, they also introduce additional complexity that must be managed by the Colonies framework. To implement a workflow management framework supporting distributed microservices, the following challenges must be addressed:

@@

\item \textbf{Fault tolerance and recovery:} In the event of executor failures, the Colonies framework must be able to re-assign processes to other executors, manage executor restarts, or trigger recovery mechanisms to maintain system reliability and resilience. In addition, the Colonies framework must be able to handle restarts or crashes of Colonies servers, thereby preventing any internal states in Colonies from becoming corrupted.

\item \textbf{Load balancing:} The Colonies framework must manage load balancing among executors to optimize performance and avoid overloading any single executor.

@@

\item \textbf{Workflow orchestration:} The Colonies framework must coordinate and orchestrate complex multi-step workflows executed by several executors, sometimes in parallel where executors run on different platforms. Colonies must define the sequence in which processes should be executed, manage dependencies among processes, and ensure that workflows execute successfully to completion.

@@

The Colonies framework uses a combination of technologies to address the aforementioned challenges. A fundamental design principle is statelessness, which makes Colonies simpler, more reliable and scalable, and easier to implement. To ensure reliability and data consistency, a distributed consensus algorithm is needed to handle high-availability and server crashes. The remainder of this section describes how the Colonies framework is implemented. The next discusses the role of queues to support batch jobs and realtime processing, but also queues roles in implementing loosely coupled systems.

@@

\subsection{The role of queues as seperation of concerns}
Separation of concerns (SoC) is a design principle to break down a complex software system into smaller, more manageable parts. For example, HTTP APIs can be used to abstract away implementation detail and provide a clear and simple interface for interacting with a particular service. However, HTTP APIs alone are insufficient for handling dynamic environments where components frequently fail or the underlying infrastructure is constantly changing. To address such environments, an alternative mechanism is needed.

@@

Queues can be implemented in different ways. While message brokers are a common solution, Colonies adopt an alternative strategy and leverage a standard database and query it for processes to assign to different executors. One key advantage of this approach is that it enables fine-grained process assignments, making it possible to assign specific processes to particular executors. For instance, an executor of the browser type can be limited to only executing processes capable of running in web browsers. This level of granularity cannot easily be implemented using message brokers which generally do not offer introspection of queues, or provide the ability to pull specific messages out of the queue. Generally, the only way to retrieve a specific message is to pull all messages from the queue, obtain the message, and then place all remaining messages back into the queue in the same order. In contrast, a database can function as a queue and a query can match any columns making it possible to assign specific executors to specific processes.

@@

When an executor connects to the Colonies server, the server hangs the incoming HTTP connection\footnote{An alternative protocol is to use WebSockets or gPRC to communicate with the Colonies server.} until the executor is assigned a process, or until a connection timer expires. The Colonies server never connects to the executors. Rather, it is the responsibility of the executors to connect to the Colonies server. This enables executors to be deployed anywhere on the Internet, behind firewalls, commercial telco networks, or even in web browser-based applications.

@@

Figure \ref{fig:process_assignment} depicts an executor submitting a function specification that is later assigned to another executor. When registering, executors have to specify to the Colonies server which functions they are capable of executing. The Colonies server then makes sure that the conditions part of the function specification matches the capability of an executor.

@@

The \emph{assign} API request binds a process from the database to an executor. Given the multi-threaded nature of the Colonies server, it is essential that the \emph{assign} request is synchronized to ensure that only one thread at a time can modify the database and update the process table, thus preventing multiple executors from being assigned to the same process. To ensure that only one executor can be assigned to a process, the \emph{assign} request must be synchronized. It is worth noting that synchronization is not necessary for other requests. For example, as the \emph{submit} request only adds new entries to the process table, there are no race conditions. The \emph{close} request sets the output of the function innovation and updates the process state to either successful or failed in the process table. Since there can only be one executor assigned to a process there are no race conditions and consequently no need for synchronization.

@@

By using the Raft protocol, Colonies can redirect incoming \emph{assign} requests to the leading Colonies server, thereby ensuring that only one Colonies server replica handles assign requests. A new leader is automatically elected in the event that a Colonies server replica fails. Figure \ref{fig:ha_deployment} shows an overview of a high-availability Colonies deployment. In addition, the Raft protocol makes it possible to seamlessly update Colonies software by making it possible to upgrade each Colonies server replica individually. Consequently, making Colonies servers well-suited for Kubernetes deployments.

@@

\subsubsection{Workflows}
A workflow is a series of processes that need to be completed in a specific order. In Colonies, workflows are represented as directed acyclic graphs (DAGs) where nodes represent processes and edges represent dependencies and data flow between processes. Similar to processes, workflows are managed completely stateless. When a workflow is submitted, all processes part of the workflow are submitted and added to the process table similar to how ordinary processes are handled. To control the order processes can be assigned, Colonies sets a flag, \emph{wait for parents} to prevent processes to be assigned to an executor before their dependencies have been satisfied. Note that processes may run in parallel if a sufficient number of executors are available, assuming their parent processes have finished.

@@

To manage data flow between processes that are part of a workflow, it is necessary to store the input and output values resulting from function invocations in a database. An example of such a database is presented in Table \ref{inouttable}. An executor can subsequently query the output from a parent process and use that as arguments while invoking a function.

@@

Cron is a time-based job scheduler designed to trigger workflows at predefined intervals. It can for example be used to automate tasks such as routinely fetching data from external systems. To make Colonies robust and easy to implement, it is essential that Cron workflows also are managed stateless. This can be achieved by introducing an additional database table to store information about Cron workflow, and let the elected Colonies server leader assume the responsibility of managing Cron workflows through a two-step process.

Firstly, the leading Colonies server calculates a future deadline, indicating when a specific Cron workflow is scheduled to be executed. The leader server then conducts periodic scans of the Cron table. When the current time surpasses the calculated deadline for a particular workflow, a new workflow is automatically submitted. Note that this protocol is completely stateless as no session information is stored in memory between cron scans. In the event that the leader Colonies server experiences a crash, a new leader is selected and assumes the responsibility of managing Cron workflows according to the two-step protocol.

@@

Generators can be implemented using a similar method to how Cron workflows are implemented. The elected leader Colonies server takes on the responsibility for scanning a special generator table to decide if a workflow should be triggered. This approach ensures that there is a single point of control for generating workflows and prevents generation of duplicate workflows. If the leader server crashes, a new leader Colonies server is automatically elected, which takes over the responsibility for triggering generators. This ensures that the generation of workflows is not impacted by server failures, and the system can continue to operate smoothly even in case of failures. It can be worth pointing out that since the \emph{pack} request only adds information to the generator table and does not manipulate any states, it can be processed by any Colonies server without synchronization precautious. This means that any Colonies server in the cluster can handle incoming \emph{pack} requests, which increases the scalability and fault-tolerance of the system as a whole.

@@

This \emph{assign} request is always initiated by the executor, allowing executors to be deployed behind firewalls. When receiving the request, the Colonies server hangs the request until a matching process is found or a timer expires. In the example in Listing \ref{code:assign}, the timer is set for 10 seconds, after which an exception is raised, and the executor must reconnect to the server.

@@

\subsection{Serverless computing}
\label{faas}
Serverless computing is a cloud-based architecture that allows developers to build and deploy applications without the need to manage the underlying infrastructure. In this model, the cloud service provider dynamically allocates and manages resources such as adding computing power or storage required by an application. Function-as-a-Service (FaaS) is a serverless computing model where developers provide code snippets that should be executed, and the cloud provider takes care of scaling, configuration and deployment of that code snippets. FaaS is often used in event-driven architectures where functions are triggered by specific events, thus eliminating the need to keep servers continuously running. The Cognit project \cite{cognit} aims to extend the FaaS concept by supporting geographically distributed edge servers in addition to cloud server resources. The project is also going to develop an AI-enabled orchestrator that can decide where to deploy FaaS code.

Colonies offer functionality to call functions in remote executors, a fundamental requirement to implement a FaaS platform. Colonies also provides mechanisms for service discovery and security isolation to prevent non-authorized function calls. In some cases, it could make sense to manually manage a colony by an administrator, and manually registering and deploying executors. Alternatively, a software solution could be developed to automatically deploy or undeploy executors. Such an approach would be very similar to a Function-as-a-Service (FaaS) platform.

Figure \ref{fig:faas} depicts a Colonies-based FaaS platform. Similar to AWS Lambda or Google Cloud Functions, developers would need to upload code snippets, which would then be packaged and published as Docker containers. To invoke a function, a function specification needs to be submitted to the Colonies server indicating which function to call.

The advantage of using Colonies to implement a FaaS platform is that it introduces an extra level of separation and the possibility to develop a loosely coupled system. For example, if an executor is missing, the process is simply enqueued until a free executor becomes available. A software agent could then monitor pending processes on the Colonies server, predict the system load, and automatically deploy executors to meet the load, resulting in a highly efficient and scalable system. In the future, the AI orchestrator developed in the Cognit project could also be used together with Colonies to reduce energy consumption, or minimize network latency by utilizing edge servers.

\subsection{Edge computing}
Edge computing is a distributed computing model that brings computing resources closer to the source where data is generated to improve efficiency or reduce latency. In edge computing, computing and processing is usually done on the \emph{edge} of a network, such as at a server connected to a 5G base station. One could for example imagine a drone video surveillance use case where parts of a video processing pipeline run on an edge server. This approach would make it possible to execute complex AI models that would otherwise be difficult to run on a drone with limited computing capabilities.

Setting up a video pipeline is a complex problem. First, the drone must be able to discover nearby edge servers, make sure the edge server has sufficient resources, deploy some software components at the edge server, then start a video server and finally connect to the video server. Orchestrating such a workflow requires configuration and provisioning of computational resources across platforms.

@@

The Colonies framework has the potential to streamline the setup of edge computing workflows by enabling devices and edge servers to be represented as distributed executors that can interact with other. Figure \ref{fig:edge} illustrates a setup of three executors. The \emph{Edge Executor} is responsible for deploying and setting up video processing software at an edge server, while the \emph{Drone Executor} is responsible for activating the video camera at the drone and establishing a connection to the video server hosted on the edge server. The \emph{Video Surveillance Executor} is responsible for identifying suitable edge servers and facilitating submission of workflows by users. Discovery of edge executors can be achieved by interacting with a Colonies server and searching the colony for available executors.

@@

A challenge associated with Slurm and HPC environments in general is that they can be difficult to set up and use. Submitting a job typically requires users to develop job scripts and manually transfer data to the HPC environment before a job can be executed. Compared to cloud platforms, Slurm is less mature when it comes to API and automation support, which can make the usage more time-consuming and manual.

@@

Colonies could be used to fill the gap between cloud and HPC and create workflows that can partly run in cloud while also taking advantage of the vaste computational capacity offered by HPC. For example, a company could develop a workflow that automatically copy training data to a HPC center, train a neural network, move the result back to the cloud, and finally re-deploy a Kubernetes service to use the new model. With Colonies, users can indirectly control HPC systems by submitting Colonies processes that are subsequently picked up by HPC executors, without requiring direct connection to the HPC system. Essentially, using Colonies as a cloud-based submission system for HPC connecting different HPC centers together. Figure \ref{fig:hpc_overlay} depicts such a submission system.

Figure \ref{fig:hpc_workflow} shows an example how Colonies could be used in a HPC environment. Several executors run at the HPC system, each with their own responsibilities. The \emph{Container Importer Executor} is responsible for converting container images to a format that is supported by Slurm, e.g. Singularity containers\footnote{https://docs.sylabs.io}. The \emph{Data Importer Executor} downloads data to the HPC file system. The \emph{Slurm Executor} is responsible for submitting and overseeing Slurm jobs, while the \emph{Data Exporter} executor is responsible for uploading result data to the cloud.

@@

\section{Discussion}
This paper has presented an open-source framework called Colonies designed to streamline execution of computational workloads across platforms. Colonies provides a \emph{distributed runtime environment}, called a \emph{Colony}, which serves as an intermediary layer to indirectly control other computer systems. Instead of directly connecting different systems together, Colonies allows services or users to publish instructions and then assign these instructions to other services. By chaining instructions together, it becomes possible to execute workloads that can operate seamlessly across different platforms.

>>>>

\begin{abstract}
Artificial intelligence and machine learning has gained significant traction in recent years. At the same time, development and operation of AI workloads has become increasingly challenging. One difficulty is lack of portability, making it cumbersome to move from one platform to another. Creating and operating fully automated end-to-end workflows across devices, edge, and cloud platforms is even more challenging.

@@

Based on a stateless microservice architecture, complex workflows can be broken down into composable functions. These composable functions can be implemented in any programming language and be executed by distributed executors deployed anywhere on the Internet such as IoT, cloud, edge, or at a supercomputer. A zero-trust security protocol enables a collection of distributed executors to operate as a unified entity, establishing seamless compute continuums across platforms.

@@

Taking an AI model into production requires substantial software engineering expertise and collaboration across teams. In contrast to traditional IT workloads, both the data and the model must be managed in addition to the software itself. As most models require regular retraining or re-calibration, it must be possible to update deployed models and software seamlessly without losing information or breaking the system. In many cases, there is a constant flow of data ingested into the system which must be managed even in case of failures. This becomes even more challenging when nodes or parts of the underlying infrastructure become unavailable due to maintenance such as software updates, hardware replacements or sometimes misconfiguration problems.

In some cases, it may be necessary to scale the system to increase or reduce the capacity. This is especially critical when using expensive cloud resources. Scaling the system means that the underlying infrastructure may change at any time, causing instability issues for running services or workflows. Therefore, it must be possible to detect failed computations and re-process failed tasks part of a workflow. Workflows must hence be designed to handle an ever-changing infrastructure, and if a failed computation cannot be restored gracefully, engineers must be able to quickly perform root cause analysis to manually recover the system.

@@

Additionally, many compute clusters operate on-premises installations. Sometimes it is necessary to temporarily increase the capacity of on-prem clusters by combining resources from multiple providers, for example, adding cloud compute resources to handle peak loads or utilize HPC resources to quickly re-process historical data. Developing hybrid workflows where some jobs run in the cloud and others run on HPC systems requires even more software development efforts \cite{wf_challenges} and is beyond the scope of many users, preventing them from utilizing powerful hardware. Clearly, there is a need for a framework that can consolidate various workflow management platforms to simplify development and enable seamless execution across platforms.

The long-term vision of the Colonies framework is to enable uninterrupted access to computing resources, allowing data to flow from one service to another establishing a seamlessly integrated computing ecosystem. To realize this vision, the paper presents \emph{Colonies}, a \emph{distributed runtime environment} enabling compute continuums. The remainder of the paper describes the Colonies framework and how it can be used to create robust and scalable AI workflows.

@@

Today, utilization of serverless computing is experiencing a significant growth \cite{cognit}, primarily due to its potential to liberate developers from the burden of managing underlying cloud infrastructures. Attempts are currently being made to implement serverless workflow management systems. For example, OpenWolf \cite{openwolf} is a serverless workflow engine designed to utilize the Function-as-a-Service (FaaS) paradigm to compose complex scientific workflows. It is based on OpenFaaS \cite{openfaas}, which allows functions to run on Kubernetes clusters.

Colonies can also be used to implement serverless workflow management systems. This will be further discussed in Section \ref{faas}. However, in contrast to previous work, Colonies is designed to be platform independent and does hence not require container orchestration platforms like Kubernetes. By using a zero-trust security protocol, functions can be securely executed by distributed executors deployed anywhere on the Internet.

Currently, microservices are primarily used to implement large-scale web applications or Internet applications requiring high-availability. It has not yet become a prevalent design principle for workload management or implementation of HPC applications. Instead, simple job scripts are commonly used. J. Represa et al. \cite{GarciaRepresa1740746, GarciaRepresa1640771} explore various challenges associated with developing microservice-based workflow management for industrial automation within the context of the Arrowhead project \cite{delsing2017iot}. The authors conclude that microservice-based workflow technologies are viable for industrial applications, particularly due to their inherent flexibility. The primary contribution of this paper is a comprehensive technical description of how to implement a distributed workflow engine based on microservice principles, extending beyond orchestrating microservices to execute automation tasks within one platform. The vision is to create a distributed architecture where microservices can reside anywhere on the Internet and still function as a cohesive unit to execute cross-platform workloads.

@@

\section{The Colonies framework}
\label{sec:headings}
A core concept of the Colonies framework is the notion of \emph{processes}. A process contains meta-information about computations executed by remote computer programs, referred to as \emph{executors}. Specifically, a process contains a definition of a function to be executed as well as contextual information such as execution status and the result of the computation. One can think of a process as a digital twin of a real computing process. It is essential to point out that a process does not necessarily have to be an operating system process, but can represent any type of computation, e.g. a remote procedure call executed by any kind of software.

@@

\emph{Executors} are responsible for executing processes assigned by the Colonies Servers. Several executors form a \emph{Colony}, which is a collection of executors operating as a cohesive unit where each executor is responsible for executing specific types of processes. One could view a Colony as an organization (like a company) of distributed computer softwares acting as a single virtual computer system. To interact with other executors, executors must prove their Colony membership using a cryptographic protocol that follows the zero-trust security principle: \emph{never trust, always verify}. This security protocol ensures that users can keep control even when executors are distributed across platforms. Zero-trust security is fundamental to the Colonies framework and will be further discussed in Section \ref{zerotrustsecurity}. The following section outlines the underlying design principles and describe the Colonies framework in more detail.

@@

\begin{itemize}
\item \textbf{Single responsibility:} Each executor is only accountable for executing specific functions. This makes the system easier to understand, develop, test and maintain.  
\item \textbf{Loosely coupled:} Executors are designed with minimal dependencies on other executors, enabling different software development teams to work independently. For instance, a data engineering team may handle the implementation of pre-processing functions, while a data science team oversees machine learning functions, and another team manages visualization or customer integration etc.

\item \textbf{Scalability:} Executors can be deployed independently and be scaled horizontally. This enables efficient resource utilization, parallelism, and improved performance.

\item \textbf{Resilience:} The failure of a single executor does not compromise the entire application or workflow. Executors' isolation from one another contributes to a more resilient and fault-tolerant system. If an executor crashes during execution, the process is automatically re-assigned to another executor.

@@

\item \textbf{Decentralized governance:} As executors are technology-agnostic, different software development teams can make independent technology and design decisions when developing new executors, promoting greater flexibility and adaptability. For example, some executors may be implemented in Rust, while others may use Python to leverage state-of-the-art machine learning frameworks.
\end{itemize}

Although microservices have attractive software engineering properties, they also introduce additional complexity that must be managed by the Colonies framework. To implement a workflow management framework supporting distributed microservices, the following challenges must be addressed:

@@

\item \textbf{Load balancing:} The Colonies framework must manage load balancing among executors to optimize the overall performance.

\item \textbf{Fault tolerance and recovery:} In the event of executor failures, the Colonies framework must be able to re-assign processes to other executors, manage executor restarts, or trigger recovery mechanisms to maintain system reliability and resilience. In addition, the Colonies framework must be able to handle re-starts or crashes of Colonies servers and prevent any internal states from becoming corrupted.

@@

\item \textbf{Workflow orchestration:} The Colonies framework must coordinate and orchestrate complex multi-step workflows executed by several executors, sometimes in parallel where executors run on different platforms. This includes defining the sequence in which processes should be executed, manage dependencies among processes, and ensure that workflows execute successfully to completion.

@@

As will be discussed, the Colonies framework uses a combination of technologies to address the aforementioned challenges. A fundamental design principle is statelessness, which makes Colonies simpler, more reliable and scalable, and easier to implement. To ensure reliability and data consistency, a distributed consensus algorithm is needed to handle high-availability and server crashes. The remainder of this section describes how the Colonies framework is implemented. The next discusses the role of queues to support batch jobs and realtime processing, but also discusses queues roles in implementing loosely coupled systems.

@@

\subsection{The role of queues as seperation of concerns}
Separation of concerns (SoC) is a design principle to break down a complex software system into smaller, more manageable parts. For example, HTTP APIs can be used to abstract away implementation detail and provide a clear and simple interface for interacting with a particular service. However, APIs alone are insufficient to handle dynamic environments where components frequently fail or the underlying infrastructure is constantly changing. To address such environments, an alternative mechanism is needed.

@@

Queues can be implemented in different ways. While message brokers are a common solution, Colonies adopt an alternative strategy and is based on a standard database. One key advantage of this approach is that it enables fine-grained process assignments, making it possible to assign specific processes to particular executors. For instance, an executor can be limited to only execute processes capable of running in web browsers. This level of granularity cannot easily be implemented using message brokers which generally do not offer introspection of queues, or provide the ability to pull specific messages out of the queue. Generally, the only way to retrieve a specific message is to pull all messages from the queue, obtain the message, and then place all remaining messages back into the queue in the same order. In contrast, a database can function as a queue and a query can match any columns making it possible to assign an executors to specific processes.

@@

When an executor connects to the Colonies server, the server hangs the incoming HTTP connection\footnote{An alternative protocol is to use WebSockets or gPRC to communicate with the Colonies server.} until the executor is assigned a process, or until a connection timer expires. The Colonies server never connects to the executors. Rather, it is the responsibility of the executors to connect to the Colonies server. This enables executors to be deployed anywhere on the Internet, behind firewalls, commercial telco networks, or even in web browsers.

@@

Figure \ref{fig:process_assignment} depicts an executor submitting a function specification that is later assigned to another executor. When registering, executors have to specify to the Colonies server which functions they are capable of executing. The Colonies server then makes sure that function specifications match the capability of executors.

@@

The \emph{assign} API request binds a process from the database to an executor. Given the multi-threaded nature of the Colonies server, it is essential that the \emph{assign} request is synchronized to ensure that only one thread at a time can modify the database and update the process table, thus preventing multiple executors from being assigned to the same process. To ensure that only one executor can be assigned to a process, the \emph{assign} request must be synchronized. It is worth noting that synchronization is not necessary for other requests. For example, as the \emph{submit} request only adds new entries to the process table, hence there are no race conditions. The \emph{close} request sets the output of the function innovation and updates the process state to either successful or failed in the process table. Since there can only be one executor assigned to a process there are no race conditions and consequently no need for synchronization.

@@

By using the Raft protocol, Colonies can redirect incoming \emph{assign} requests to the Colonies server replica elected as the leader, thereby ensuring that only one Colonies server replica handles assign requests. A new leader is automatically elected in the event that a Colonies server replica fails. Figure \ref{fig:ha_deployment} shows an overview of a high-availability Colonies deployment.

@@

\subsubsection{Workflows}
A workflow is a series of processes that need to be completed in a specific order. In Colonies, workflows are represented as directed acyclic graphs (DAGs) where nodes represent processes and edges represent dependencies and data flow between processes. Like processes, workflows are managed completely stateless. When a workflow is submitted, all processes part of the workflow are submitted and added to the process table similar to how ordinary processes are handled. To control the order processes are assigned, Colonies sets a flag, \emph{wait for parents} to prevent processes to be assigned to an executor before their dependencies have been satisfied. Note that processes may run in parallel if a sufficient number of executors are available, assuming their parent processes have finished.

@@

To manage data flow between processes that are part of a workflow, it is necessary to store the input and output values from function executions in the database. An example of such a database is presented in Table \ref{inouttable}. An executor can subsequently query the output from a parent process and use that as arguments while invoking a function.

@@

Cron is a time-based job scheduler designed to trigger workflows at predefined intervals. It can for example be used to automate tasks such as routinely fetching data from external systems. To make Colonies robust and easy to implement, it is essential that Cron workflows also are managed statelessly. This can be achieved by introducing an additional database table to store information about Cron workflow, and let the elected Colonies server replica leader assume the responsibility of managing Cron workflows through a two-step process.

Firstly, the Colonies server replica leader calculates a future deadline, indicating when a specific Cron workflow is scheduled to be executed. The leader server then conducts periodic scans of the Cron table. When the current time surpasses the calculated deadline for a particular workflow, a new workflow is automatically submitted. Note that this protocol is completely stateless as no session information is stored in memory between cron scans. In the event that the leader Colonies server crashes, a new leader is selected and assumes the responsibility of managing Cron workflows according to this two-step protocol.

@@

Generators is implemented using a similar method like how Cron workflows are implemented. The elected leader Colonies server takes on the responsibility for scanning a special generator table to decide if a workflow should be triggered. This approach ensures that there is a single point of control for generating workflows and prevents generation of duplicate workflows. If the leader server crashes, a new leader Colonies server is automatically elected, which takes over the responsibility for triggering generators. This ensures that the generation of workflows is not impacted by server failures, and the system can continue to operate smoothly even in case of failures. It can be worth pointing out that since the \emph{pack} request only adds information to the generator table and does not manipulate any states, it can be processed by any Colonies server without synchronization precautious. This means that any Colonies server in the cluster can handle incoming \emph{pack} requests, which increases the scalability and fault-tolerance of the system as a whole.

@@

This \emph{assign} request is always initiated by the executors, allowing executors to be deployed behind firewalls. When receiving the request, the Colonies server hangs the request until a matching process is found or a timer expires. In the example in Listing \ref{code:assign}, the timer is set for 10 seconds, after which an exception is raised, and the executor must reconnect to the server.

@@

\subsection{Serverless computing}
\label{faas}
Serverless computing is a cloud-based architecture that allows developers to build and deploy applications without the need to manage the underlying infrastructure. In this model, the cloud service provider dynamically allocates and manages resources such as adding computing power or storage required by an application. Function-as-a-Service (FaaS) is a serverless computing model where developers provide code snippets that should be executed, and the cloud provider takes care of scaling and deployment of that code snippets. FaaS is often used in event-driven architectures where functions are triggered by specific events, thus eliminating the need to keep servers continuously running. The Cognit project \cite{cognit} aims to extend the FaaS concept by supporting geographically distributed edge servers in addition to cloud server resources. The project is also going to develop an AI-enabled orchestrator that can decide where to deploy FaaS code.

Colonies offer functionality to call remote functions, a fundamental requirement to implement a FaaS platform. Colonies also provides mechanisms for service discovery and security isolation to prevent non-authorized function calls. A software solution could be developed to automatically deploy and undeploy executors. Such an approach would be very similar to a Function-as-a-Service (FaaS) solution.

Figure \ref{fig:faas} depicts a Colonies-based FaaS platform. Similar to AWS Lambda or Google Cloud Functions, developers need to upload executors source code, which would then be automatically packaged and published as Docker containers. To invoke a function, a function specification needs to be submitted to the Colonies server indicating which function to call.

The advantage of using Colonies to implement a FaaS platform is that it introduces an extra level of separation and the possibility to develop a loosely coupled system. For example, if an executor is missing, the process is simply enqueued until a auitable executor becomes available. A software agent could then monitor pending processes on the Colonies server, predict the system load, and automatically deploy executors to meet the load, resulting in a highly efficient and scalable system. In the future, the AI orchestrator developed in the Cognit project could also be used together with Colonies to reduce energy consumption or minimize network latency.

\subsection{Edge computing}
Edge computing is a distributed computing model that brings computing resources closer to the source where data is generated to improve efficiency or reduce latency. In edge computing, computing and processing is usually done on the \emph{edge} of a network. One could for example imagine a drone video surveillance use case where parts of a video processing pipeline run on an edge server at 5G base station. This approach would make it possible to execute complex AI models that would otherwise be difficult to run on a drone with limited computing capabilities.

Setting up a video pipeline utilizing edge resources is a complex problem. First, the drone must be able to discover nearby edge servers, make sure the edge server has sufficient resources, deploy some software components at the edge server, then start a video server and finally connect to the video server. Orchestrating such a workflow requires configuration and provisioning of computational resources across platforms.

@@

The Colonies framework has the potential to streamline the setup of edge computing workflows by enabling devices and edge servers to be represented as executors that can interact with other. Figure \ref{fig:edge} illustrates a setup of three executors. The \emph{Edge Executor} is responsible for deploying and setting up video processing software at an edge server, while the \emph{Drone Executor} is responsible for activating the video camera at the drone and establishing a connection to the video server hosted on the edge server. The \emph{Video Surveillance Executor} is responsible for identifying suitable edge servers and facilitating submission of workflows by users. Discovery of edge executors can be achieved by interacting with a Colonies server and searching the colony for available executors.

@@

A challenge associated with Slurm and HPC environments in general is that they can be difficult to set up and use. Submitting a job typically requires users to develop job scripts and manually transfer data to the HPC environment before a job can be executed. Compared to cloud platforms, Slurm is less mature when it comes to API and automation support, which can make the use of HPC environments more time-consuming.

@@

Colonies could be used to fill the gap between cloud and HPC and create workflows that can partly run in cloud while also taking advantage of the vaste computational capacity offered by HPC. For example, a company could develop a workflow that automatically copy training data to a HPC center, train a neural network, move the result back to the cloud, and finally re-deploy a Kubernetes service to use the new model. With Colonies, users can indirectly control HPC systems by submitting Colonies processes that are subsequently picked up by HPC executors, without requiring direct connection to the HPC system. Essentially, using Colonies as a cloud-based submission system for HPC to connect several HPC centers together. Figure \ref{fig:hpc_overlay} depicts such a submission system.

Figure \ref{fig:hpc_workflow} shows an example how Colonies could be used in a HPC environment. Several executors run at the HPC system, each with their own responsibilities. The \emph{Container Importer Executor} is responsible for converting container images to a format that is supported by Slurm, e.g. converting Docker containers to Singularity containers\footnote{https://docs.sylabs.io}. The \emph{Data Importer Executor} downloads data to the HPC file system. The \emph{Slurm Executor} is responsible for submitting and overseeing Slurm jobs, while the \emph{Data Exporter} executor is responsible for uploading result data to the cloud.

@@

\section{Discussion}
This paper has presented an open-source framework called Colonies providing a \emph{distributed runtime environment} designed to streamline execution of computational workloads across platforms. Instead of directly connecting different systems together, Colonies allows special kind of microservices so-called executors to publish instructions which are subsequently assigned to other executors. By chaining instructions together, it becomes possible to execute workloads that can operate seamlessly across different platforms.
